Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-07-25 Paper/intro.tex
..

Time evolution of a chemically reacting system is largely dependent upon rate constants
associated with individual reactions. The rate constants are typically assumed to exhibit
a certain correlation with temperature (e.g. Arrhenius-type). Hence, accurate specification
of the rate-controlling parameters is critical to the fidelity of simulations. However,
in practical applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data. Consequently, the
underlying uncertainty associated with the parameters and therefore the simulation-prediction
is typically ignored. Vohra et al. calibrated the Arrhenius diffusion parameters to quantify
the uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian framework
in~\cite{Vohra:2014} and~\cite{Vohra:2017}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments. Bayesian calibration
was accelerated in these efforts using a polynomial chaos~\cite{Ghanem:2003, Xiu:2002} 
surrogate for the model. However,
the number of uncertain parameters for calibration were limited to two in both cases. In
higher dimensions, the construction of surrogate as well as the calibration process become
remarkably challenging due to increase in computational effort.

In this paper, we focus our efforts on mitigating the computational challenge posed by uncertainty
quantification (UQ) for applications involving a large number of chemical reactions. Specifically, we
accomplish this by implementing an efficient approach that aims to reduce the dimensionality
of the problem by computing the so-called active subspace~\cite{Constantine:2015}. 
The active subspace, discussed later in~\ref{sub:ac}, essentially constitutes the dominant
eigenspace of a matrix derived from the gradient information of the model output with respect
to an input. The high-dimensional application problem considered in this work is the H$_2$/O$_2$
reaction mechanism~\cite{Yetter:1991}. This mechanism has been gaining a lot of attention in the
recent years as a potential source of clean energy for locomotive applications~\cite{Das:1996}.   
The mechanism involves 19 reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species:

@@

For each reaction in table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.  The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
%
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
%
The equivalence ration ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer (O$_2$) ratio at given
condition to the same quantity under stoichiometric conditions. In this study, computations were performed
at fuel-rich conditions, $\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of temperature increase
to exceed a given threshold, regarded as `ignition delay' is recorded.

@@

The dominant eigenvectors that constitute the active subspace correspond to the important directions in the
parameter space. In other words, the variability in the model output is predominantly observed along these
directions. Moreover, the components of these eigenvectors can be used to estimate the so-called 
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into the relative importance
of the uncertain parameters. In 
section~\ref{sub:gsa}, we demonstrate that the activity scores could be used to approximate the 
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and establish a relationship between
the Sobol' total effect index, the DGSMs, and the activity scores.

>>>>

Time evolution of a chemically reacting system is largely dependent upon rate constants
associated with individual reactions. The rate constants are typically assumed to exhibit
a certain correlation with temperature (e.g. Arrhenius-type). Hence, accurate specification
of the rate-controlling parameters is critical to the fidelity of simulations. However,
in practical applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data. Consequently, the
underlying uncertainty associated with the parameters and therefore the simulation-prediction
is typically ignored. Vohra et al. calibrated the Arrhenius diffusion parameters to quantify
the uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian framework
in~\cite{Vohra:2014}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments. Additionally, 
in~\cite{Vohra:2017}, a design strategy was implemented to determine optimal conditions
for experiments in order to enhance the information gain in the uncertain diffusion parameters.
Bayesian calibration was accelerated in these efforts using a polynomial 
chaos~\cite{Ghanem:2003, Xiu:2002} surrogate for the model. However,
the number of uncertain parameters for calibration were limited to two in both cases. In
higher dimensions, the construction of surrogate as well as the calibration process become
remarkably challenging due to increase in computational effort.

In this paper, we focus our efforts on mitigating the computational challenge posed by uncertainty
quantification (UQ) for applications involving a large number of chemical reactions. Specifically, we
accomplish this by implementing an efficient approach that aims to reduce the dimensionality
of the problem by computing the so-called active subspace~\cite{Constantine:2015}. 
The active subspace, discussed later in~\ref{sub:ac}, essentially constitutes the dominant
eigenspace of a matrix derived from the gradient information of the model output with respect
to an input. The dominant eigenspace essentially comprises `important' directions in the parameter
space that predominantly capture the variability in the model output due to uncertainty in the
inputs.

The high-dimensional application problem considered in this work is the H$_2$/O$_2$
reaction mechanism~\cite{Yetter:1991}. This mechanism has been gaining a lot of attention in the
recent years as a potential source of clean energy for locomotive applications~\cite{Das:1996}.   
The mechanism involves 19 reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in table~\ref{tab:kinetics}, 
the reaction rate is assumed to show an Arrhenius correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
%
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
%
The equivalence ration ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer (O$_2$) ratio at given
condition to the same quantity under stoichiometric conditions. In this study, computations were performed
at fuel-rich conditions, $\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of temperature increase
to exceed a given threshold, regarded as `ignition delay' is recorded.

@@

Components of the dominant eigenvectors can be used to estimate the so-called 
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into the relative importance
of the uncertain parameters. In 
section~\ref{sub:gsa}, we demonstrate that the activity scores could be used to approximate the 
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and establish a relationship between
the Sobol' total effect index, the DGSMs, and the activity scores. Below, we summarize key contributions of
the paper and its organization.

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity scores as means for approximating
the DGSMs is provided. (2) A general relationship between DGSMs, activity scores, and the total Sobol' index 
is established. (3) Iterative gradient-based and gradient-free approaches for computing the active subspace are
 demonstrated. The two approaches are shown to be consistent and yield computational gains compared to the 
 conventional approach. As mentioned above, the gains are greater in the case of grad-free approach. (4) The
 grad-free approach is implemented for discovering an active subspace in a 33-dimensional parameter space 
 for the H$_2$/O$_2$ reaction kinetics application.

 \paragraph{Organization}
 The paper is organized as follows. In Section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In Section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In Section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the grad-based, and the grad-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the grad-free approach. In Section~\ref{sec:app},
we further explore the suitability of the grad-free approach by implementing it to a higher-dimensional
H$_2$/O$_2$ reaction kinetics problem by accounting for the uncertainty in $E_{a,i}$'s as well. 
Finally, we draw important conclusions based on the various aspects of this work in Section~\ref{sec:conc}.
